#1.How big is the dataset?
```{r}
dim(gapminder)
```
#2.How many variables are in the dataset? What are they?
```{r}
colnames(gapminder)
```
#3.How would you examine the first and last few rows of the dataset
```{r}
head(gapminder)
tail(gapminder, n=10)
```
#4.What is the class of each of the variables? In other words, what types of variables are they?
```{r}
str(gapminder)
```
#5.How many countries are represented
```{r}
length(unique(gapminder$country))
```
#6.How many continents are represented?
```{r}
length(unique(gapminder$continent))
```
#7.Create a new dataset containing only the data for Australia
```{r}
onlyAus <- subset(gapminder, gapminder$country=="Australia")
```
#8.Calculate the increase in average life expectancy for Australians over the time period in the dataset.
```{r}
tail(onlyAus$lifeExp, n=1) - head(onlyAus$lifeExp, n=1)
```
#9.Produce pairwise scatterplots of the three numerical variables for Australia.
```{r}
#produce pairwise, scallterplots
plot(onlyAus[4:6])
plot(subset(onlyAus, select =c(lifeExp, pop, gdpPercap)), labels =c("life expectancy", "population", "GDP per capita"))
```
#10.What is the increase in life expectancy for Burundi?
```{r}
burundilifeexp <- subset(gapminder, gapminder$country=="Burundi")

tail(burundilifeexp$lifeExp, n=1) - head(burundilifeexp$lifeExp, n=1)
```
#11.In which rows of the dataset does Swaziland occur?
```{r}
which(gapminder$country=="Swaziland")
```
#12. How would you create a dataset that contains only the data for 2007 for all countries?
```{r}
Only2007 <-   subset(gapminder, gapminder$year=="2007")
Only2007
```
#13.Which country had the lowest life expectancy in 2007? Which had the highest?
```{r}
Only2007[which.min(Only2007$lifeExp),]
Only2007[which.max(Only2007$lifeExp),]
```
#14. What is the mean GDP per capita in each country?
```{r}
tapply(gapminder$gdpPercap, gapminder$country, mean)
```
#15.Do all countries have data in the same range of years?
```{r}
tapply(gapminder$year, gapminder$country, range)
```
#16.Construct a dataset consisting of only the life expectancy over time for all countries.
```{r}
LifeExp <- gapminder$lifeExp
LifeExp <- matrix(LifeExp, nrow = 12, ncol = 142)
colnames(LifeExp) <- unique(gapminder$country)
rownames(LifeExp) <- unique(gapminder$year)
LifeExp[1:12, 1:142]
```
#17.Plot a time series plot of the life expectancy against time for all countries on the same plot.
```{r}

FiveCols <-c("red3", "royalblue4", "springgreen4", "mediumpurple3", "orange2")
ContinentCols <-FiveCols <-c("red3", "royalblue4", "springgreen4", "mediumpurple3", "orange2")[tapply(gapminder$continent,gapminder$country, 
                                              function(x){
                                                unique(x)
                                              })]
matplot(rownames(LifeExp), LifeExp, xlab="Year", ylab="Life Expectancy", type = "l", lty = 1, col = ContinentCols, ylim = c(10,82))
legend("bottomright", col=FiveCols, legend=levels(gapminder$continent),lty=1, bty="n")
```






























