```{r}
library(gapminder)
```
#1. Construct a plot of the average life expectancy over the last four years in each country against its average GDP per capita over that same time period, and label it appropriately.
```{r}
le <- gapminder$lifeExp
le <- matrix(le,nrow=12, byrow = FALSE)
le <- le[9:12,]
avgle4 <- colMeans(le)

gdp <- gapminder$gdpPercap
gdp <- matrix(gdp, nrow=12, byrow = FALSE)
gdp <- gdp[9:12,]
avggdp4 <- colMeans(gdp)
plot(avggdp4, avgle4, xlab="GDP per capita", ylab = "Life expectancy")
```



#2. What do you notice about the scale of GDP? Transform it appropriately and re-plot.
```{r}
plot(avggdp4, avgle4, xlab="GDP per capita", ylab = "Life expectancy", log = "x")
```

#3. Use different symbols for countries in different continents. Hint: you will need the arguments pch. so see the help file for plot. First create a vector that gives you the continent that each country is in.
```{r}
#gapminder$continent[!duplicated(gapminder$country)]
#or
Continent <- gapminder$continent[!duplicated(gapminder$country)]

plot(avggdp4, avgle4, xlab="GDP per capita", ylab = "Life expectancy",log="x", pch=as.numeric(Continent))
```

#4. Add a legend so that we know which symbol corresponds to which continent.
```{r}
plot(avggdp4, avgle4, xlab="GDP per capita", ylab = "Life expectancy",log="x", pch=as.numeric(Continent))
legend("bottomright", legend=sort(unique(Continent)), pch = 1:5)
```


#5. Construct a plot similar to Q4, but this time use the same symbol but different colours for eachcontinent.
```{r}
plot(avggdp4, avgle4, xlab="GDP per capita", ylab = "Life expectancy",log="x",  col=as.numeric(Continent))
legend("bottomright", legend=sort(unique(Continent)),col =1:5 , pch=1)
```

