---
title: "week5"
author: "Kai Tang"
date: "3/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


 

#2. What is the size of these two files?
```{r}
dim(MovieFinances)
dim(IMDB)
```

#3. Most operating systems will have utilities that will allow you to determine the number of lines in a file (and other information as well) without having to open the file in a program. Google how to do this at the Windows command prompt.

   <!-- 
   wc -l <filename> prints the line count (note that if the last line does not have \n, it will not be counted)
    wc -c <filename> prints the byte count
    wc -m <filename> prints the character count
    wc -L <filename> prints the length of longest line (GNU extension)
    wc -w <filename> prints the word count
    -->
    
#4 Before importing a file into R, it’s useful to know something about its structure. Again, use Google to find out how to view the contents of a text file at the command line.

#5. Using the command read.csv, read in the data from these two files to create two data frames. For the purposes of this workshop, call them IMDB and MovieFinances. The values of some of the arguments will depend on the structure of the file that you saw in 3. above.
```{r}
MovieFinances <- read.csv("C:/Users/zzzta/Desktop/data-science/Workshop 5/MovieFinances.csv", header = TRUE)
IMDB <- read.csv("C:/Users/zzzta/Desktop/data-science/Workshop 5/IMDB.csv", header = TRUE)
```

#6. After importing the .csv files into R, have a look at the variable names in both data frames, and then decide whether you need to modify some of them so that they are more compact or more meaningful.
```{r}
colnames(IMDB)
#colnames(MovieFinances)
# "Budget..M."          "Domestic.Gross..M." "Worldwide.Gross..M."  varilabile is not user friendly
colnames(MovieFinances[c(5,6,7)] <- c("Budget", "DomesticGross", "WorldwideGross"))
colnames(MovieFinances)

```


#7. Are there any variables that both datasets have in common? Which ones? Are there any superfluous columns/variables? If there are, remove them.

Common -Release.Year and Year, Movie and Title.
Superfluous - (IMDB -- Var1, year, budget)
```{r}
#IMDB <- IMDB[,1]
#IMDB <- IMDB[]
IMDB = subset(IMDB , select = -c(var1))
```
#8. Use head to look at the first few rows of each data frame, or view them in the RStudio data viewer. Do you notice anything that might be unusual?
```{r}
#some data is not available
```

#9. Examine the structure of the variables in each data frame. What do you notice about the type of the variables title and mpaa in IMDB and Movie and Month in MovieFinances? What should we do about them?

```{r}
levels
```
# 10. So it looks like we’re going to do need to do some manipulation, but perhaps it’s better to merge the data sets together because they’ll only have some of the same common elements. Have a look at the help file for the function merge, and then merge the two datasets together. Call the result AllData. What variables should we merge on?
```{r}
AllData <- merge(IMDB, MovieFinances, by.x = c("title", "year"), by.y = c("Movie", "Release.Year"))
```

#11.  How large is the merged dataset?
```{r}
dim(AllData)
nrow(AllData)
ncol(AllData)
```
#12. Try out the function summary using the data frame as the argument. What kind of information does it produce? What unusual aspects do you notice?
```{r}
summary(AllData)

#summary for Action, Ant etc is not really helpful
```
#13. Unfortunately, summary doesn’t really tell us about missing values, but the function describe in the package Hmisc does. Install the package from the Tools menu, and then load the library Hmisc to be able to use describe. Scan the output of describe and try to understand what it’s telling you.
```{r}
require(Hmisc)
describe(AllData)
```
#14. The function describe also has a nice feature: if you save the results of describe into an object, and then plot that object, you’ll get a couple of plots that might be useful. What do those plots tell you about the distributions of the variables?
```{r}
desc <- describe(AllData)
plot(desc)
```
#15 Note that there are two columns with budget information. what are the characteristics of those columns? Do they give the same information? What plot could you construct to determine whether they do?
```{r}
par(pty="s")
plot(Budget ~ I(budget/10^6), data = AllData, xlab = "budget/1M")
```

